---
title: "Elegant Color in R Advanced Graphics"
date: 2021-02-13
author: "***Qianjiang Hu***"
output: 
  html_document:
      number_sections: false
      toc: true
      toc_depth: 4
      toc_float:
        collapsed: true
        smooth_scroll: true
---

## **Overview**
R is a programming language and free software environment for statistical computing and graphics supported by the R Foundation for Statistical Computing. It is also extremely flexible with high perfermance in creating data visualisations. Above all, the color scheme is one of the key elements of a data visualization.
This post will present colors in basic R and ggplot.

## **Colors in Basic R**
### Built-in color names
R has 657 built in color names. That can be listed by `colors()`. 
```{r}
colors()[c(1:10)]
```

Instead of using a color name, color can also be called by Hex code. Hex code define a color as a 6 hexadecimal digit number of the form `#RRGGBB`, in which `RR` is for red, `GG` for Green and `BB` for blue. Two additional characters (with the same scale) can be added to the end to describe transparency (#rrggbbaa). The function `seecol()` in `unikn` package offers a way to view color palettes. You can search a perfect color by an online tool `color picker`. <https://htmlcolorcodes.com/>
```{r, echo=TRUE, eval=TRUE}
library('unikn')             # loads the package
seecol(c('#3338FF','#33FF48','#FF338D','#AC33FF','#FF6033',"#FF000044"))
```
### RGB color via `rgb()`
- Combination of numbers for red, green and blue component
- Default range between 0 and 1
- maxColorValue = 255 adjusts the range to the standard notation
- alpha produces transparent colors

```{r, fig.width=12, fig.height=6, fig.align='center', out.width= '80%'}
par(mar = c(7,3,2,2))
barplot(1:5, col =
          c(rgb(red = 1, green = 0, blue = 0, alpha = 0.5),  # alpha produce the transparent color
            rgb(red = 1, green = 1, blue = 1), 
            rgb(red = 0.5, green = 0.5, blue = 0.5), 
            rgb(0 , 90, 160, maxColorValue = 255),
            rgb(140, 180, 35, maxColorValue = 255)), 
        names.arg = 
          c('rgb(1,0,0)','rgb(1,1,1)','rgb(0.5,0.5,0.5)','rgb(0,90,160)','rgb(140,180,35)'), 
        las = 2)
```
R can translate colors to rgb by `col2rgb()` (It gives interaction for matching colors in other programs)
```{r}
col2rgb(c('#FF0000', 'blue'))
```
### HCL color
Package `colorspace` support the mapping HCL(Hue-Chroma-Luminance) color system, a perceptually-based and device-independent color model.
- `qualitative_hcl()` create colors with same perceptual weight
- `sequential_hcl()` create colors of decreasing luminance
- `diverging_hcl()` create transmissions between two colors
- `palette()` presents pre-specified patterns
- `n` sets number of steps
- `h`, `c`, `l`: parameters of color coding alternative
- use `hcl_palettes(plot = TRUE)` to check all examples
- `power` determines the speed of change: `power = 1`: linear, `power = 2`: quadratic

```{r, fig.width=8, fig.height=4, fig.align='center', out.width= '80%'}
library(colorspace)
col1 <- sequential_hcl(n = 7, palette = "Blues 3")
col2 <- diverging_hcl(n = 7, palette = "Blue-Red2")
col3 <- qualitative_hcl(7, palette = "Dark 3")
seecol(col1)
seecol(col2)
seecol(col3)
```
Customized color verctors are also available with colorspaceby vector.
```{r, fig.width=8, fig.height=4, fig.align='center', out.width= '80%'}
col4 <- sequential_hcl(n = 7, palette = "Blues 3")[c(1,3,5)]
seecol(col4)
```

There is also an interactive tool that can be used to obtain a customized palette. To start the tool:
`colp <- choose_palette()`

### ColorRampPalette
Another powerful color schemes is ColorRampalette with both `colorRamp()` and `colorRampPalette()` function. With two input colors, both functions can generate an entire spectrum of colors by mixing together different amounts of two colors.
```{r}
cr <- colorRamp(c("red", "blue"))
cr(0)
``` 
Any value between 0 and 1, even a sequence of numbers, can be passed to colorRmp.

```{r}
cr <- colorRamp(c("red", "blue"))
cr(seq(0,1,len = 8))
``` 
The `colorRampPalette()` shares the similar funtion with `colorRamp()`, however the function that it returns gives you a fixed number of colors that interpolate the palette. 
```{r}
cr <- colorRampPalette(c("red", "blue"))
cr(1)
cr(3)
``` 

`colorRampPalette(colors = c(.,.), alpha)(n = ...)` build sequential and divering color-schemes based on rgb coding. Same color for start and end with different transparency produce sequential scheme. Two different colors for start and end produces diverging scheme.

```{r, fig.width=8, fig.height=3, fig.align='center', out.width= '80%'}
par(mfrow = c(1,2))
barplot(1:6, main = "Diverging", 
        col = colorRampPalette(colors = c(rgb(0, 0, 1), rgb(0, 1, 0)))(n = 6))
barplot(1:6, main = "Sequential", 
        col = colorRampPalette(colors = c(rgb(0, 0, 1, alpha = 1),
                                          rgb(0, 0, 1, alpha = 0)),
                               alpha = TRUE)(n = 6))
```

In addition, `colorRamp()` generates contrast color palette, while `colorRampPalette()` generates a gradient color pallette. 


```{r, fig.width=10, fig.height=5, fig.align='center', out.width= '80%'}
par(mfrow = c(1, 2))
pc1 <- colorRamp(c("blue", "purple"))
pc2 <- colorRampPalette(c("blue", "purple"))
pie(rep(1, 20), labels = "", col = pc1(seq(0, 1, len = 20)), 
    border = pc1(seq(0, 1, len = 20)), main = "colorRamp")
pie(rep(1, 20), labels = "", col = pc2(20), border = pc2(20), 
    main = "colorRampPalette")
dev.off()
```

### RColorBrewer
The function of `brewer.pal()` in `RColorBrewer` package support three type of pallettes: `RColorBrewer`, `Diverging`,`Qualitative`. 
All of these palettes can be used in conjunction with the `colorRamp()` and `colorRampPalette()`.
`display.brewer.all()` can show all available color pallettes from `RColorBrewer` package.


```{r, fig.width=8, fig.height=8, fig.align='center', out.width= '90%'}
library(RColorBrewer)
display.brewer.all()
```

```{r}
library(RColorBrewer)
cols <- brewer.pal(3, "BuGn")
cols
```

```{r, fig.width=7, fig.height=5, fig.align='center', out.width= '80%'}
pal <- colorRampPalette(cols)
image(volcano, col = pal(20))
```

### Other Palettes
There are also some recommended palettes built-in R:
`cm.colors()`, `topo.colors()`, `terrain.colors()`, `heat.colors()`, `rainbow()`, `gray.colors()`


## **Colors in `gglot2`**

Most of the color palettes including `R names` and `Hex code` are still available in `ggplot2`. Customized color vectors are also work in `ggplot2`. Additionally, ggplot2 is compatible with some other color palettes.

### fill vs. col
`ggplot2` allows to customize the colors thanks to its `fill` and `color` arguments. In general, `fill` give the color with which a geom is filled, whereas `color` define the colors with which a geom is outlined.

```{r, fig.width=7, fig.height=5, fig.align='center', out.width= '80%'}
library(ggplot2)
ggplot(data = iris, mapping = aes(x = Sepal.Length, y =Sepal.Width)) +
  geom_point(col = 'red', fill = 'deepskyblue', shape = 21, size = 5, alpha = 0.7)
```

### Default colors in ggplot2
Default colors are defined by `hcl()` with the discipline `hcl(h = (seq(15, 375-360/n, length = n)), c = 100, l = 65)`. Therefore, all colors have the same luminance and chroma.
Additionally, `show_col()` in packages `scales` offer the other way to check colors.
```{r, fig.width=7, fig.height=4, fig.align='center'}
# Visualize the colors depending on the number of categories by
par(mfrow = c(1,2))
library(scales)
show_col(hue_pal() (4))    #please note the space between `hue_pal()` and `n`
show_col(hue_pal() (6))
```

### Set a ggplot color by groups 
(i.e. by a factor variable)
This is done by mapping a grouping variable (aesthetic) to the color or to the fill arguments. 

```{r, fig.width=7, fig.height=3, fig.align='center', out.width= '80%'}
library('ggplot2')
sleep <- as.data.frame(msleep)
Plot1 <- ggplot(data = sleep, aes(x = vore)) +
  geom_bar(aes(fill = vore), col = "black") # fill define the aesthetic mapping in which variable `vore` will be filed with color
Plot1
```

### Discrete colors
#### *`scale_fill_hue`*
Change colors in ggplot2 by `scale_fill_hue` and `scale_color_hue`
Use grey scale via `scale_fill_grey()`
`na.value`: Change color of category `NA`

```{r, fig.width=7, fig.height=3, fig.align='center', out.width= '80%'}
par(mfrow = c(1,2))
Plot1 +
  scale_fill_hue(l = 100)
Plot1 +
  scale_fill_grey()
```

#### *`cale_fill_brewer()`*
`Color brewer` also offer a function to change colors in R: `scale_fill_brewer()` and `scale_color_brewer()`
Consider missing values via `na.value`, otherwise the category is not displayed.
To reverse the order of the colors use the argument direction = -1

```{r, fig.width=7, fig.height=3, fig.align='center', out.width= '80%'}
Plot1 + 
  scale_fill_brewer('Paired')
```

Check all colors in `RcolorBrewer` via `RColorBrewer::display.brewer.all(n)`

#### *`scale_fill_manual()`*
Define discrete colors manually in ggplot2 via `scale_fill_manual()` and `scale_color_manual()`
Firstly, it is necesary to build either a vector of colors or a vector of categories and colors.

```{r, fig.width=7, fig.height=3, fig.align='center', out.width= '80%'}
newcol <- c("carni" = "lightblue", "herbi" = "lightgreen", "insecti" = "tomato", "omni" = "purple")
Plot1 + 
  scale_fill_manual(values = newcol, na.value = "grey")
```

Similar with `scale_fill_manual()`, `scale_shape_manual()` giva a way to self-define shapes of categories via a shape vector.

```{r, fig.width=7, fig.height=3, fig.align='center', out.width= '80%'}
mcol <- c('setosa' = 'lightblue',  'versicolor'= 'lightgreen', 'virginica' = 'tomato')
mshape <- c('setosa' = 22,  'versicolor'= 23, 'virginica' = 24)
ggplot(data = iris,  mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(fill = Species, shape = Species), size =  4, alpha = 0.75)+
  scale_fill_manual(values = mcol, na.value = 'grey')+
  scale_shape_manual(values = mshape, na.value = 9)
```

### Gradient Colors

When a categrory variable is used to asethetic, discrtete colors are required. Whereas, when a continuouse variable is used for scaling, then a gradient color is necessary.

```{r, fig.width=7, fig.height=4, fig.align='center', out.width= '80%'}
Plot2 <- ggplot(data = iris,  mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length, size = Petal.Width),  alpha = 0.7)
Plot2
```

* `scale_fill_gradient()` creates a two color gradient (low-high)
* `scale_fill_gradient2()` creates a three color gradient (low-mid-high, midpoint)

```{r, fig.width=7, fig.height=4, fig.align='center', out.width= '80%'}
par(mfrow = c(1,2))
Plot2 + scale_color_gradient(low = 'white', high = 'blue')
Plot2 + scale_color_gradient2(low = 'blue', mid = 'white',high = 'red', midpoint = 4)
```

#### ***limits of continuous colors***

The upper and lower end of the gradient color scheme can be changed via the argument `limits`
• scale_fill_gradient() changes also the color pattern
• scale_fill_continuous() changes only the limits

```{r, fig.width=7, fig.height=4, fig.align='center', out.width= '80%'}
par(mfrow = c(1,2))
Plot2 + scale_color_continuous(limits= c(-2,10))
Plot2 + scale_color_gradient2(low = 'blue', mid = 'white',high = 'red', midpoint = 4, limits = c(0,9))
```

### Global default of colors
`options()` Allow the user to set and examine a variety of global options which affect the way in which R computes and displays its results.

Global adjustment of color pattern via options(), e.g.
* options(ggplot2.continuous.colour = brewer.pal(8, "Dark2"))
* options(ggplot2.continuous.fill = brewer.pal(8, "Dark2"))
* options(ggplot2.discrete.colour = brewer.pal(8, "Dark2"))
* options(ggplot2.discrete.fill = brewer.pal(8, "Dark2"))

***NOTE***
Before changing colors globally, save the default via:

```{r}
opts <- options()
```

## **Others**

### Color info
To extract colors (other info as well) of a ggpplot use `ggplot_build()`

```{r}
ggplot_build(Plot1)$data[[1]]
```






## **References:**
1. <https://htmlcolorcodes.com/>
2. <https://www.r-graph-gallery.com/ggplot2-color.html>
3. <https://bookdown.org/rdpeng/exdata/plotting-and-color-in-r.html>
4. <https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf>





