---
title: "r batch install and load packages"
date: 2022-03-31
author: "***Qianjiang Hu***"
output: 
  html_document:
      number_sections: false
      toc: true
      toc_depth: 4
      toc_float:
        collapsed: true
        smooth_scroll: true
---

## **General installation methods**
- CRAN Mirror 

|        `install.packages('package_name')`
|        - `repo = "https://lib.ugent.be/CRAN/"` to indicate respiratory
|        - `dependencies = TRUE` install all of the dependencies
|        > `getCRANmirrors()` show available mirrors
|            
| 
    
- BiocManager 

|        `if (!requireNamespace("BiocManager")) {install.packages("BiocManager")}` 

|        `BiocManager::install('package_name')` 
|          
| 
 
- devtools after `install.packages("devtools"`) 

|        - `install_bioc()` from Bioconductor,
|        - `install_bitbucket()` from Bitbucket,
|        - `install_cran()` from CRAN,
|        - `install_git()` from a git repository,
|        - `install_github()` from GitHub,
|        - `install_local()` from a local file,
|        - `install_svn()` from a SVN repository,
|        - `install_url() `from a URL, and
|        - `install_version()` from a specific version of a CRAN package.
|        - e.g. `devtools::install_github("packages")`
|          
| 
 
## **General methods to load packages**

- library 

|        `library('package_name')`
|            
| 

- require 

|        `require('package_name')`
|            
| 

- call function via `::`  

|        `package::function()`
|         - `::` is quite helpful if there are namespace collisions; e.g. functions from different packages with the same name.
| 

## **Install multiple packages**


```{r, echo=TRUE, eval= FALSE}
rm(list=ls()) # clean r space and objects
```


```{r, echo=TRUE, eval= FALSE}
list_of_packages <- c('DESeq2',
                      'ggplot2',
                      'ggfortify',
                      'biomaRt',
                     'org.Mm.eg.db',
                      'tidyverse',
                      'ComplexHeatmap',
                      'cowplot',
                      'circlize',
                      'RColorBrewer',
                     'clusterProfiler',
                     'enrichplot',
                      'msigdbr',
                     'DOSE',
                     'dplyr',
                    'EnhancedVolcano',
                     'pathview',
                     'fgsea',
                    'ggnewscale'
                     )

new_packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,'Package'])]

if(length(new_packages) >0) BiocManager::install(new_packages, dependencies=TRUE)

```


## **Load multiple packages**


```{r, echo=TRUE, eval= FALSE}

sapply(list_of_packages, library, character.only = T)

```


## **Set working directory**
```{r, echo=TRUE, eval=FALSE}

getwd()
here::here() # get current working folder path or the path of specific files

### e.g. Set your working directory
here::here()
cm_dir <- here::here('count_matrix')
rs_dir <- here::here('results')
pl_dir <- here::here('plots')
sp_dir <- here::here('scripts')
me_dir <- here::here('metadata')


### get file path and sample ID

files <- dir('count_matrix', pattern = "*.count")   # get all file path with the suffix of '.count' under the folder of 'count_matrix'
file_ID <- unlist(strsplit(files, split = '.count'))   # extract the file name and set as sample ID
file_ID <- gsub(pattern = '.count', replacement = '', files) ## alternative; grep the files names and replace the '.count' as ''.
## sub() and gsub(): The functions which are exclusively useful for string substitution operations in R. You can replace the string in a vector or a data frame and can substitute the specified string.

```



















